Time Series Analysis of Chaotic Dynamical Systems with Recurrent Neural Networks

### 2019-04-12 doc/paper/paper.tex
numerical methods section written

@@

\subsection{History}

@@

% TODO: talk about attractors

@@

Now, the $L^2$ norm of error is 
\begin{align*}
    \sqrt{(x_\epsilon - x)^2 + (y_\epsilon - y)^2} &= \sqrt{(r_\epsilon - r)^2\sin^2(t) + (r_\epsilon - r)^2\cos^2(t)} \\
    &= \sqrt{(r_\epsilon - r)^2(\sin^2(t) + \cos^2(t))^2} \\
    &= \sqrt{(r_\epsilon - r)^2} \\
    &= \vert r_\epsilon - r \vert.
\end{align*}

@@

Empirically shown, \echonn can be used to identify the max LCE of smaller 
dynamical systems, and special precaution has to be taken when choosing a 
time interval. Using a larger time will yield a more accurate $\lambda_1$, but 
it may also cause an overflow if the system is chaotic. Thus, 
maximum time values need to be chosen such that $\vert \vert \Mv\vert 
\vert_{\infty}$ doesn't exceed $10^{100}$ (where $\vert \vert \cdot \vert 
\vert_{\infty}$ is the $L^\infty$ norm).

@@

% TODO: numerical solutions section

@@

The Lorenz system is named after its creator, but it is also known as the
Lorenz butterfly due to the shape of the attractor of the system. The Lorenz
Butterfly can be seen in Figure \ref{fig:lorenz_butter}.

@@

and where mass body 1 and two are located at $-\alpha$ and $\mu$.

@@

%%% MODELS %%%
\section{Models}

\subsection{ARIMA}
%\subsection{Recurrent Neural Networks}
\subsection{Echo State Networks}
%\subsection{LTSM}

>>>>

Neural Networks are known to be able to approximate any function given 
enough data and time to train \cite{cybenko1989approximation}.
In light of this, the hope is that a deep learning based model,
the ESN, will outperform ARIMA.

@@

%\subsection{History}

@@

@@

Now, the $\ell^2$ norm of error is 
\begin{align*}
    \sqrt{(x_\epsilon - x)^2 + (y_\epsilon - y)^2} &= \sqrt{(r_\epsilon - r)^2\sin^2(t) + (r_\epsilon - r)^2\cos^2(t)} \\
    &= \sqrt{(r_\epsilon - r)^2(\sin^2(t) + \cos^2(t))^2} \\
    &= \sqrt{(r_\epsilon - r)^2} \\
    &= \vert r_\epsilon - r \vert.
\end{align*}

@@

Empirically shown, \echonn can be used to identify the max LCE of smaller 
dynamical systems, and special precaution has to be taken when choosing a 
time interval. Using a larger time will yield a more accurate $\lambda_1$, but 
it may also cause an overflow if the system is chaotic. Thus, 
maximum time values need to be chosen such that $\vert \vert \Mv\vert 
\vert_{\infty}$ doesn't exceed $10^{100}$ (where $\vert \vert \cdot \vert 
\vert_{\infty}$ is the $\ell^\infty$ norm).

@@

\subsection{Numerical Integration Methods}

In order to generate the data to be analyized, the solutions to these 
dynamical systems need to be computed. Each of the systems are defined by 
a system of differential equations which can be numerically solved. 
However, due to the chaotic nature, error introduced by the numerical 
solution will grow exponentially, so precaution need to be taken to avoid 
having this sort of error introduced into the collected data. When 
numerically solving chaotic problems, it's possible for the chaotic 
behavior to be suppressed, or where there shouldn't be chaotic behavior,
it could be created \cite{corless1994good}.

However, it's been shown that the Fourth Order Runge-Kutta Method (RK4)
outperforms other numerical methods such as Euler's Method and the
Midpoint Method for chaotic systems \cite{zidan2011effect}.
Not only does RK4 perform better than other methods, it serves as a good
method for producing accurate numerical solutions to choatic systems with 
small enough time steps: the Lü chaotic system \cite{mehdi2017using} and
the Zhou system \cite{roslan2013solving}, for example. One article points
out that a larger concern 
regarding these solutions is the accuracy of the initial conditions 
measured from the physical problem which the mathematical model is based,
but otherwise, they echo the strength of correctly implemented 
numerical solutions \cite{corless1994good}. Further, my initial conditions
are randomly generated, so this is not a concern.

I will be using an improved version of RK4 which
incorporates the fifth order Runge-Kutta: RK4(5)
\cite{dormand1980family}. This is the method implemented in
the Python function \texttt{scipy.integrate.solve\_ivp}, and my
numerical solver
wraps this function to integrate the chaotic systems. Using small time
steps will yield accurate samples from the chaotic systems.

@@

The Lorenz system is named after its creator, but it is also known as the
Lorenz butterfly due to the shape of the system. The Lorenz
Butterfly can be seen in Figure \ref{fig:lorenz_butter}.

@@

and where mass body one and two are located at $-\alpha$ and $\mu$.

@@

\section{The Echo State Network}

\newcommand{\Tv}{\mathbf{T}}
\newcommand{\Uv}{\mathbf{U}}
\newcommand{\Dv}{\mathbf{D}}
\newcommand{\tv}{\mathbf{t}}
\newcommand{\uv}{\mathbf{u}}
\newcommand{\dv}{\mathbf{d}}
\newcommand{\yv}{\mathbf{y}}
\newcommand{\Fv}{\mathbf{F}}
\newcommand{\wv}{\mathbf{w}}

The Echo State Network (ESN) is a type of Recurrent Neural Network (RNN)
used for time series analysis, but it is trained differently than most
RNNs and is 
characterized by having the echo state property. I implement the design
detailed by Jaeger in is 2002 book \cite{jaeger2002tutorial} and
further detailed in his 2007 article \cite{jaeger2007echo}.

A time series data set will contain samples of data which change
over time. Part of the data at each time point is measured or known
before, called input data and denoted by $\uv(t)$ with $K$ dimensions.
The rest of the data will be the goal for prediction, called output data and denoted $\dv(t)$ with $L$ dimensions.

For $T$ time samples, $t_1, t_2, \dots, t_T$, we have a $T \times K$ 
matrix, $\Uv$, with sample input vectors as rows. We also
have a $T \times L$ matrix,
$\Dv$, with sample output vectors as rows. A model will predict the 
values of $\dv$, and this prediction will be denoted $\yv$.

\subsection{Recurrent Neural Networks and Traditional Training Method}
\label{sec:rnn}
RNNs are usually designed by having $K$ input units, $L$ output units, 
and some internal units organized by some network structure. These units
have evolving values for every time point. The
structure of networks describes how units are connected. 
The input units at time $t_i$ are set to $\uv(t_i)$. However, output
units and
internal units are calculated from the values of other units. These other 
units are said to be connected to the resultant unit. Output
units are calculated from current input units, current internal units,
and from previous output units. Internal units 
are calculated from the values of the current input units, 
previous internal units, and previous output units.

Although, it's rare that the output units are calculated from previous 
output units \cite{jaeger2002tutorial}. Sometimes for training, the output 
units are set to the value of $dv$, called teacher forcing, so that the 
model trains on actual values as opposed to its accumulated errors 
\cite{jaeger2002tutorial}.

To calculate the value of a unit, an activation function is applied to a 
linear combination of the units connected to it \cite{svozil1997introduction}. The constants for the 
linear combination are called weights, and they are parameters of the 
model. Which activation function is used is also a parameter of the 
function. Typically, some non-linear function is used, such as $\tanh$, 
sigmoid, or rectified linear unit.

During training, we search for
weights which maximize the model's performance on data
which the model hasn't seen. A common training method for RNNs
is to use a form of backpropagation \cite{jaeger2002tutorial}. 
Backpropagation for RNNs is slightly different from traditional Neural Networks, but the idea is the same.

For example, given some neural network model $\yv(\xv; \wv)$, where
$\xv$ is known input data, $\wv$ is the model's parameters, and $\yv$ 
approximates some values, $\tv$, then we define an objective 
function as the sum of squared error:

\newcommand{\Ev}{\mathbf{E}}

\begin{equation}
    \Ev(\xv; \wv) = \vert \vert \yv(\xv; \wv) - \tv \vert \vert_2^2.
    \label{eq:error_eq_rnn}
\end{equation}

Using steepest descent to solve this minimization problem (as shown in 
\cite{svozil1997introduction}), we can
reduce $\Ev$ in (\ref{eq:error_eq_rnn}) by
updating the weights as follows:

\begin{equation}
    \wv = \wv - \lambda \nabla_\wv \Ev
\end{equation}

for some learning rate $\lambda$.

\subsection{Echo State Network Implementation}

The implementation design used for this paper is detailed by Herbert Jaeger
in \cite{jaeger2002tutorial} and \cite{jaeger2007echo}. I explain it here 
to clarify my exact process (and with a slight change of notation to be 
consistent with the rest of this paper).

Like RNNs, ESNs have an input layer with $K$ units and with values 
$\uv(t_1), \uv(t_2), \dots,  \uv(t_T)$ where $\uv(t) \in \mathbb{R}^K$.
$K$ is zero if there are no input units. There is an internal layer
with $N$ units, denoted by $\xv(t_i) \in \mathbb{R}^N$ for
$i = 1, 2, \dots, T$, and 
there is an output 
layer with $L$ units, denoted by $\yv(t_i) \in \mathbb{R}^L$ for all
$i = 1, 2, \dots, T$.

\newcommand{\W}{\mathbf{W}}
\newcommand{\Win}{\mathbf{W}_{in}}
\newcommand{\Wback}{\mathbf{W}_{back}}
\newcommand{\Wout}{\mathbf{W}_{out}}

How the units are connected can vary, but the suggested way 
\cite{jaeger2002tutorial} is to connect the internal layer to the current 
input layer, the previous internal layer, and the previous output layer. 
These connections will have weights described in \ref{sec:rnn}. 
Respectively, these weights will be denoted by the matrices:
$\Win$, $\W$, and $\Wback$. The activation function for the internal
layer is $\tanh$.

The output layer is connected to the current input layer and current 
internal layer. The weight for these connections will be denoted $\Wout$ 
(we only have one matrix since the connected layers are concatenated into a
single vector, per Jaeger's notation \cite{jaeger2002tutorial}). The 
activation function for these connections is linear. Although, the 
recommended choice
is $\tanh$, the output layer doesn't necessarily lie in the range of 
$\tanh$, so it can't be used without some form of normalization.

In mathematical notation, the model is defined as follows:

\begin{align}
    \xv(t_{i}) &= \tanh \left(
            \Win \uv(t_i)
            + \W \xv(t_{i-1})
            + \Wback \yv(t_{i-1})
        \right) \\
    \yv(t_i) &= \Wout (\uv(t_i), \xv(t_i))_{\text{concat}}.
\end{align}

$\xv$ and $\yv$ are calculated from previous unit values but for the 
first time step, there are none, so we need to initialize the
units to zeros \cite{jaeger2002tutorial}. 
This is done by defining $\xv(t_0)$ and $\yv(t_0)$ to zeros.

The weights are initialized such that the model has the defining 
characteristic of echo state networks: the echo state property
(Section \ref{sec:eprop}).
During the training phase (Section \ref{sec:etrain}), $\Wout$ is fit to
the training data.

Once the network has been initialized and trained, it can be used to 
perform time series analysis.

\subsection{The Echo State Property}
\label{sec:eprop}

The echo state property is defined by the following \cite{jaeger2002tutorial}:

\begin{quote}
    Assume an untrained network with weights $\Win$, $\W$, and $\Wback$ is 
    driven by teacher input $\uv(n)$ and teacher-forced by teacher output 
    $\dv(n)$ from compact intervals $\Uv$ and $\Dv$. The network ($\Win$, 
    $\W$, and $\Wback$) has echo states w.r.t. $\Uv$ and $\Dv$, if for 
    every left-infinite input/output sequence $(\uv(n), \dv(n-1))$, where 
    $n=\dots, -2, -1, 0$, and for all state sequences $\xv(n)$,
    $\xv'(n)$ compatible with the teacher sequence, i.e. with
    \begin{align}
        \xv(n) &= \tanh \left(
                \Win \uv(n)
                + \W \xv(n-1)
                + \Wback \yv(n-1)
            \right) \\
        \xv'(n) &= \tanh \left(
                \Win \uv(n)
                + \W \xv'(n-1)
                + \Wback \yv(n-1)
            \right)
    \end{align}
    it holds that $\xv(n) = \xv'(n)$ for all $n \leq 0$.
\end{quote}

The idea is that if the network is run long enough, it will, in a sense, 
forget long past input and focus only on the most recent input. If two 
networks with differing internal states are run with the same input and 
teacher-forced output, they will eventually converge to each other.
It's name suggests data fed to the model will echo around but eventually 
fade as echos do.

Jaeger outlines a process to create weights which will likely have
echo state properties. The process is posed as conjecture and 
doesn't guarantee the echo state property. However, the echo state is very 
likely and can be
assumed confidently \cite{jaeger2002tutorial} \cite{jaeger2007echo}.

To demonstrate the echo state property, initialize the network with random 
values. Set user input (if any) and teacher-forced output units to zero,
After enough iterations, the internal state should converge to zero.

The process to initialize the echo state weights, $\W$, is as follows \cite{jaeger2002tutorial}:

\newcommand{\specrad}{\vert \lambda_\text{max} \vert}
\begin{enumerate}
    \item Initialize an $N \times N$ matrix, $\W_0$, with values from a 
    uniform distribution in $[-1, 1]$.
    \item Calculate the spectral radius, $\specrad$, of $\W_0$, i.e., the 
    maximum absolute value of the eigenvalues of $\W_0$.
    \item Normalize $\W_0$ to $\W_1$ so that it has unit spectral radius:
    $\W_1 = \frac{1}{\specrad} \W_0$
    \item Choose parameter $\alpha < 1$. Typically, $\alpha$ lies in
    the range $(0.7, 0.98)$ \cite{jaeger2002tutorial}. This number will 
    represent the new spectral radius of $\W$.
    \item Scale $\W_1$ to $\W_2$ to have a spectral radius of
    $\alpha$: $\W_2 = \alpha \W_1$. Now, let $\W = \W_2$, and the model
    will have the echo state property \cite{jaeger2002tutorial}.
\end{enumerate}

If the spectral radius of $\W$ is found to be greater than 1, then it can 
be shown that the model doesn't demonstrate the echo state property; 
Jaeger's process uses this fact by ensuring that this isn't the case 
\cite{jaeger2002tutorial}. While it can't guarantee an echo state property,
the resultant matrix does have the property frequently enough that it's a
reliable process. In his words, the model "has always been found to be an 
echo state network" \cite{jaeger2002tutorial}.

\newcommand{\vv}{\mathbf{v}}

Applying linear algebra, this makes a lot of sense. Let $A$ be an
$N \times N$ matrix with spectral radius $r = \rho(A) > 0$ and denote its
eigenvalues $\lambda_i$ and eigenvectors $\vv_i$ for
$i = 1, 2, \dots, N$.  Then, define $B = \frac{\alpha}{r} A$.

By definition, $\vv_i A = \lambda_i \vv_i$, so 
\begin{align*}
   \vv_i B &= \frac{\alpha}{r} \vv_i A \\
   &= \frac{\alpha}{r} \lambda_i \vv_i
\end{align*}

Therefore, $B$ has eigenvalues $\frac{\alpha}{r} \lambda_i$ for all $i$. 
Further, its spectral radius is then

\begin{align*}
    \vert\vert \frac{\alpha}{r} \lambda \vert\vert_\infty
    &= \frac{\alpha}{r} \vert\vert \lambda \vert\vert_\infty \\
    &= \frac{\alpha}{r} r \\
    &= \alpha
\end{align*}

where $\lambda$ is a vector of eigvenvalues and
$\vert \vert \cdot \vert \vert_\text{max}$ is the $\ell^\infty$ norm.

As the model iterates, the matrix transformation of $B$ is repeatedly 
applied. With $\uv$ and teacher-forced $\dv$ as zeros, the internal state
has the value $\xv(t_{k+1}) = \tanh(B \xv_k)$ for iteration $k$.

Since the spectral radius of $B$ is less than 1, 
\begin{align*}
    \lim_{k \to \infty}B^k \vv_i &= \lim_{k \to \infty} \lambda_i^k \vv_i & \text{ (where $\lambda_i$ and $\vv_i$ is an eigvenvalue-eigenvector pair of $B$)} \\
    &= 0 \vv_i & \text{ (since $\vert\lambda_i\vert < \alpha < 1$ and $\lim_{k \to \infty}\alpha^k = 0$)} \\
    &= 0
\end{align*}

This suggests that as the model iterates, the internal state converges to 
zero.

The rest of weights can be initialized however desired, so we will use a 
common technique.

\begin{enumerate}
    \item Initialize an $N \times M$ matrix, $\W_0$, with values from a 
    uniform distribution in $[-1, 1]$. $M$ and $N$ here are not the 
    same values from before for the sake for generality. 
    \item Scale the matrix by $\frac{1}{\sqrt{M}}$ to calculate the weight matrix.
\end{enumerate}

Then, $\Win$ and $\Wback$ can be calculated as follows:

\begin{align}
    \Win &= \frac{\W_0}{\sqrt{K}} \nonumber \\
    \Wback &= \frac{\W_0}{\sqrt{L}} \nonumber
\end{align}

Different from most training techniques, $\Win$, $\W$, and $\Wback$ are
never changed after initialization.

\subsection{Echo State Network Training}
\label{sec:etrain}

In order to train the echo state network, we need to find good 
weight values for $\Wout$. Training the echo state network is different
from the backpropagation approach in that we can solve for
and calculate directly the weights matrix; more modern methods use
an iterative approach which converges to good weight values. Part of the
reason is due to the calculation cost of the iterative methods. Today, that
it is no longer an issue, but historically, it was a strong quality of
the echo state network \cite{jaeger2007echo}.

First, the data (input, $\uv$, and output, $\dv$, values) is partitioned into three groups:
\begin{enumerate}
    \item Initialization values at time points
    $t_0, t_1, \dots, t_{T_0-1}$.
    \item Training values from
    $t_{T_0}, \dots, t_T$
    \item Testing values from
    $t_{T+1}, \dots, t_{T_f}$
\end{enumerate}

Second, the initialization data is fed into the model. This initializes the
inner state of the model, and the initial state (zeros) is lost due to the 
echo state property. $T_0$ should be chosen such that this is truly the 
case. Like $\alpha$, $T_0$ is a parameter of the echo state network and
differs depending on the problem. Good values for $T_0$ are 10 to 500 
depending on the problem complexity.

Third, the training values are fed into the model to sample the network.
We can store the input unit values, the internal state values, and the 
desired output values as matrices $M$ and $T$. $M$ will store the input 
unit values and internal state value pairs
$(\uv(t_{T_0}), \xv(t_{T_0})), \dots, (\uv(t_T), \xv(t_T))$
in its rows. $T$ will store the output unit values
$\dv(t_{T_0}), \dots, \dv(t_T)$.

To train the model, we want $\Wout M$ to approximate accurately
$g^{-1}(T)$, where $g$ is the activation function for the output layer.
Since we use an idempotent output layer activation function,
$g^{-1}(T) = T$.

Therefore, $\Wout$ should be found such that $\Wout M = T$ 
\cite{jaeger2002tutorial}.
Solving for $\Wout$,

\begin{align*}
    \Wout M &= T \\
    (\Wout M)^T &= T^T & \text{ (where $A^T$ is the transpose of $A$)} \\
    M^T \Wout^T &= T^T \\
    (M^T)^{-1} M^T \Wout^T &= (M^T)^{-1} T^T & \text{ (where $A^{-1}$ is the Moore-Penrose pseudoinverse \cite{ben2003generalized} of $A$)} \\
    \Wout^T &= (M^T)^{-1} T^T \\
    \Wout &= \left((M^T)^{-1} T^T\right)^T \\
    \Wout &= T \left((M^T)^{-1}\right)^T
\end{align*}

Using packages like \texttt{numpy.linalg.eig} and 
\texttt{numpy.linalg.pinv}, solving for eigenvalues and
pseudoinverses for $\Wout$ is very 
easy. After solving for $\Wout$, the echo state network is trained.

To evaluate the network, the testing values are fed into the model. 
However, this time, the output layer is not teacher-forced since 
in real application, the output values are unknown. 
To measure how well the echo state network approximates the desired output
we define the error to be the root mean squared error (RMSE) of the
actual output values $\dv$ and
predicted values $\yv$: $\sqrt{\sum_{i=T+1}^{n} (\dv(t_i) - \yv(t_i))^2}$ 
where $n \in [T+1 \dots T_f]$. The error represents the predictive power
of the model for $n$ time steps in the future. If RMSE is low for larger 
values of $n$, the model has a strong long term predictive power.

%noise: 0.0001 to 0.01 \cite{jaeger2002tutorial}

%\subsection{LTSM}
